{\bf {\Large Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico}}

### 2016-09-28 ident_manu.Rnw
inits and strcs supp tables added

A set of heuristics was developed to balance the tradeoff in model complexity and identifiability given the challenges described above \citep[see also][]{Wagener01}.  These rulesets were developed with the assumption that parameters will be selected with preference for those with high sensitivity and identifability based on $\gamma < 15$ as an acceptable threshold for subsets (e.g., 93\% accountability).  Selection heurestics also recognized that parameter categories (i.e., optics, organic matter, phytoplankton, temperature, zooplankton) may have unequal preferences by model users given questions of interest.  In all selection scenarios, parameters were selected by decreasing sensitivity starting with the most sensitive until identifiability did not exceed $\gamma = 15$ where selections were \begin{inparaenum}[1\upshape)]
\item blocked within parameter category,
\item independent of parameter category,
\item or considering all categories equally.
\end{inparaenum} The selection rules produced seven subsets of parameters that could further be used to optimize model calibration for \ac{do}.

@@

The effects of observational and structural uncertainty on the sensitivity analyses were evaluated for \ac{do}.  These analyses used changes in parameter uncertainty and identifiability from results using the model's default configuration to infer the effects of observational and structural uncertainty on model output.  First, observational uncertainty was evaluated by varying the initial conditions supplied to the model.  The initial conditions were based on observational data from research cruises in the northern \ac{gom} \citep{Murrell14} and uncertainty in these data translates directly to uncertainty that can influence results of the sensitivity analysis.  For example, the sensitivity of \ac{do} to variation in the half-saturation constants for phytoplankton (the concentration supporting half the maximum uptake rate of nutrients) will vary given the initial nutrient concentrations \citep{Eppley69}.  Further, changes in the ratio between nitrogen and phosphorus could affect sensitivity depending on which nutrient is limiting.  Parameter sensitivity and identifiability was re-evaluated for \ac{do} by varying all initial conditions that were non-zero by 50\% and 200\% of the original values that were estimated from the observed data.  Initial conditions were also changed to test the effects of different limiting nutrients by setting nitrogen and phosphorus values above or below concentrations defined by the standard Redfield ratio.

The effects of model structure on parameter sensitivity and identifiability were evaluated by including or excluding specific components of the model.  The FishTank model includes several processes that are based on expected conditions or available data.  These `switches' are conceptually different from model parameters as they define the use of explicit equations or processes included in the model structure.  As such, a comparison of sensitivity and identifiability given different equations uses parameter uncertainty to begin addressing componenents of structural uncertainty.  Switches in FishTank include different structural equations for the vertical attenuation of light through the water column (inherent or apparent optical properties, \citealt{Penta09,Eldridge10}) and chlorophyll to carbon ratio models (fixed or dynamic given light and nutrients, \citealt{Cloern95}).  Several switches also affect phytoplankton growth including different models for specific growth and effects of temperature, light dependence, nutrient uptake, and internal cell quotas \citepalias[][references therein]{LehrterIR}.  For simplicity, parameter identifiability was evaluated using three scenarios of simple, intermediate, and complex combinations of model switches.

@@

The above analyses were repeated for additional state variables estimated by FishTank to provide further descriptions of ecological dynamics that are relevant for hypoxia.  In addition to \ac{do}, other state variables included \ac{chla}, irradiance, nitrate, ammonium, particulate organic matter, dissolved organic matter, and phosphorus.  Particulate and dissolved organic matter were estimated as the summation of the respective outputs for organic matter from phytoplankon (\textit{OM1\_A}, \textit{OM2\_A}), fecal pellets (\textit{OM1\_fp}, \textit{OM2\_fp}), river sources \textit{OM1\_rp}, \textit{OM2\_rp}), and boundary conditions (\textit{OM1\_bc}, \textit{OM2\_bc}, see \citetalias{LehrterIR}).

@@

>>>>

A set of heuristics was developed that address the tradeoff in model complexity and identifiability given the challenges described above \citep[see also][]{Wagener01}.  These rulesets were developed with the assumption that parameters will be selected with preference for those with high sensitivity and identifability based on $\gamma < 15$ as an acceptable threshold for subsets (e.g., 93\% accountability).  Selection heurestics also recognized that parameter categories (i.e., optics, organic matter, phytoplankton, temperature, zooplankton) may have unequal preferences by model users given questions of interest.  In all selection scenarios, parameters were selected by decreasing sensitivity starting with the most sensitive until identifiability did not exceed $\gamma = 15$ where selections were \begin{inparaenum}[1\upshape)]
\item blocked within parameter category,
\item independent of parameter category,
\item or considering all categories equally.
\end{inparaenum} The selection rules produced seven subsets of parameters that could further be used to optimize model calibration.

@@

The effects of observational and structural uncertainty on the sensitivity analyses were evaluated by changing the initial conditions and structural components, respectively, from the default model setup.  First, observational uncertainty was evaluated by varying the initial conditions that were based on observational data from research cruises in the northern \ac{gom} \citep{Murrell14}. Uncertainty in these data translates directly to uncertainty that can influence results of the sensitivity analysis.  For example, the sensitivity of \ac{do} to variation in the half-saturation constants for phytoplankton (the concentration supporting half the maximum uptake rate of nutrients) will vary given the initial nutrient concentrations \citep{Eppley69}.  Further, changes in the ratio between nitrogen and phosphorus could affect sensitivity depending on the limiting nutrient.  Parameter sensitivity was re-evaluated by varying all initial conditions that were non-zero by different seasonal means based on averages of water quality data across stations and years.  April and September seasonal averages of the observed data were used to evaluate the effects of conditions that were typical of spring and late summer on the \ac{lcs}.

The effects of model structure on parameter sensitivity were evaluated by changing specific components of the model.  The FishTank model includes several ecosystem processes or characteristics that can be included based on expected conditions, available data, or desired complexity.  These `switches' are conceptually different from model parameters as they allow the inclusion of explicit equations or processes.  Switches in FishTank include different structural equations for the vertical attenuation of light through the water column (inherent or apparent optical properties, \citealt{Penta09,Eldridge10}) and chlorophyll to carbon ratio models (fixed or dynamic given light and nutrients, \citealt{Cloern95}).  Several switches also affect phytoplankton growth including different models for specific growth and effects of temperature, light dependence, nutrient uptake, and internal cell quotas \citepalias[][references therein]{LehrterIR}.  Parameter sensitivity was evaluated by comparing the results from the default model setup to an alternative, more complex, setup that included additional processes with added switches.

@@

The above analyses were repeated for additional state variables estimated by FishTank to provide further descriptions of ecological dynamics that are relevant for hypoxia.  In addition to \ac{do}, other state variables that were evaluated were \ac{chla}, irradiance, nitrate, ammonium, particulate organic matter, dissolved organic matter, and phosphorus.  Particulate and dissolved organic matter were estimated as the summation of the respective outputs for organic matter from phytoplankon (\textit{OM1\_A}, \textit{OM2\_A}), fecal pellets (\textit{OM1\_fp}, \textit{OM2\_fp}), river sources \textit{OM1\_rp}, \textit{OM2\_rp}), and boundary conditions (\textit{OM1\_bc}, \textit{OM2\_bc}, see \citetalias{LehrterIR}).

@@

% initial conditions evaluated 
<<inits, eval = T, cache = F>>=
data(ini_eval)

inits <- lapply(ini_eval, function(x) unlist(x[[1]])) %>% 
  do.call('rbind', .) %>% 
  t %>% 
  data.frame(vars = row.names(.), .) %>% 
  mutate(
    Default = sapply(def, scinot),
    April = sapply(apr, scinot), 
    September = sapply(sep, scinot)
  ) %>% 
  select(-def, -apr, -sep) %>% 
  mutate(
    vars = factor(vars, 
      labels = c('Ammonium', 'Dissolved Inorganic Carbon', 'Dissolved Oygen','Nitrate', 'Phosphate', 'Silica'),
      levels = c('NH4', 'DIC', 'O2', 'NO3', 'PO4', 'Si')
    ), 
    vars = as.character(vars)
  )

cap_val <- 'Initial conditions that were varied to evaluate effects of observational uncertainty on parameter sensitivity.  April and September values are based on seasonal averages from water quality samples of the \\ac{lcs}.  All units are $\\mu$mol L$^{-1}$.'

latex( 
  inits[, -1],
  file = '',
  rowlabel = 'Variables',
  caption = cap_val,
  caption.loc = 'top',
  label = 'tab:inits',
  rowname = inits[, 1]
  )  
@

% structural conditions
<<strcs, eval = T, cache = F>>=
swtch <- data.frame(
  lab = c('Temperature', 'Uptake', 'Quota', 'Chla:C', 'Photosynthesis', 'Growth'),
  sim = c('sigmoidal', 'Michaelis-Menten', 'Droop', 'regression', 'photoinhibition', 'minimum'),
  com = c('Arrenhius', 'Geider', 'Flynn', 'Cloern', 'nutrient dependent', '\\textit{umax} is nutrient dependent'), 
  stringsAsFactors = F
  ) %>% 
  rename(
    `Switch type` = lab, 
    Default = sim, 
    Complex = com
  )

cap_val <- 'Model switches that were used to evaluate effects of structural uncertainty on parameter sensitivity.'

latex( 
  swtch[, -1],
  file = '',
  rowlabel = 'Switch type',
  caption = cap_val,
  caption.loc = 'top',
  label = 'tab:strcs',
  rowname = swtch[, 1]
  )

@
\clearpage

### 2016-09-28 ident_manu.tex
inits and strcs supp tables added

\begin{singlespace}
\title{{\bf {\Large Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, John C. Lehrter$^1$, Lisa L. Lowe$^2$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:lehrter.john@epa.gov}{lehrter.john@epa.gov}}}
	\\\\{\textit {\normalsize $^2$Lockheed Martin IS \& GS - Civil, supporting the USEPA}}
	\\{\textit {\normalsize Resarch Triangle Park, NC 27709}}
	\\{\textit {\normalsize Phone: 919-541-3985,}}
	\\{\textit {\normalsize Email: \href{mailto:lowe.lisa@epa.gov}{lowe.lisa@epa.gov}}}
  \vspace{1in} 
  \\ Version Date:   Wed Sep 28 08:25:08 2016 -0500
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

>>>>

\begin{singlespace}
\title{{\bf {\Large Parameter sensitivity and identifiability for a biogeochemical model of hypoxia in the northern {G}ulf of {M}exico}}}
\author{
  {\bf {\normalsize Marcus W. Beck$^1$, John C. Lehrter$^1$, Lisa L. Lowe$^2$}}
  \\\\{\textit {\normalsize $^1$USEPA National Health and Environmental Effects Research Laboratory}}
  \\{\textit {\normalsize Gulf Ecology Division, 1 Sabine Island Drive, Gulf Breeze, FL 32561}}
	\\{\textit {\normalsize Phone: 850-934-2480, Fax: 850-934-2401}}
	\\{\textit {\normalsize Emails: \href{mailto:beck.marcus@epa.gov}{beck.marcus@epa.gov}, \href{mailto:lehrter.john@epa.gov}{lehrter.john@epa.gov}}}
	\\\\{\textit {\normalsize $^2$Lockheed Martin IS \& GS - Civil, supporting the USEPA}}
	\\{\textit {\normalsize Resarch Triangle Park, NC 27709}}
	\\{\textit {\normalsize Phone: 919-541-3985,}}
	\\{\textit {\normalsize Email: \href{mailto:lowe.lisa@epa.gov}{lowe.lisa@epa.gov}}}
  \vspace{1in} 
  \\ Version Date:   Wed Sep 28 10:02:54 2016 -0500
	}
\date{}
\maketitle
\end{singlespace}
\clearpage

@@

% initial conditions evaluated 
%latex.default(inits[, -1], file = "", rowlabel = "Variables",     caption = cap_val, caption.loc = "top", label = "tab:inits",     rowname = inits[, 1])%
\begin{table}[!tbp]
\caption{Initial conditions that were varied to evaluate effects of observational uncertainty on parameter sensitivity.  April and September values are based on seasonal averages from water quality samples of the \ac{lcs}.  All units are $\mu$mol L$^{-1}$.\label{tab:inits}} 
\begin{center}
\begin{tabular}{llll}
\hline\hline
\multicolumn{1}{l}{Variables}&\multicolumn{1}{c}{Default}&\multicolumn{1}{c}{April}&\multicolumn{1}{c}{September}\tabularnewline
\hline
Dissolved Inorganic Carbon&$2134$&$2096.5$&$2133.37$\tabularnewline
Ammonium&$1.09$&$1.79$&$0.35$\tabularnewline
Nitrate&$71.4$&$3.75$&$2.04$\tabularnewline
Dissolved Oygen&$172$&$194.47$&$152.57$\tabularnewline
Phosphate&$1.81$&$0.23$&$0.42$\tabularnewline
Silica&$71.4$&$5.84$&$9.77$\tabularnewline
\hline
\end{tabular}\end{center}

% structural conditions
%latex.default(swtch[, -1], file = "", rowlabel = "Switch type",     caption = cap_val, caption.loc = "top", label = "tab:strcs",     rowname = swtch[, 1])%
\begin{table}[!tbp]
\caption{Model switches that were used to evaluate effects of structural uncertainty on parameter sensitivity.\label{tab:strcs}} 
\begin{center}
\begin{tabular}{lll}
\hline\hline
\multicolumn{1}{l}{Switch type}&\multicolumn{1}{c}{Default}&\multicolumn{1}{c}{Complex}\tabularnewline
\hline
Temperature&sigmoidal&Arrenhius\tabularnewline
Uptake&Michaelis-Menten&Geider\tabularnewline
Quota&Droop&Flynn\tabularnewline
Chla:C&regression&Cloern\tabularnewline
Photosynthesis&photoinhibition&nutrient dependent\tabularnewline
Growth&minimum&\textit{umax} is nutrient dependent\tabularnewline
\hline
\end{tabular}\end{center}
